Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-01 paper.qmd
name of bib file standardised
correct kruskal refence included

bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
---

@@

<!-- - Connection between NLDR and MDS-->
Historically, \kD{}   representations of \pD{} data have been computed using multidimensional scaling (MDS) [@kruskal1964], which includes principal components analysis (PCA) [@jolliffe2011] as a special case. (@borg2005 provides a comprehensive guide to MDS.) The \kD{}   representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{}   layout is constructed by minimizing a stress function that differences distances between points in \pD{}   with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

@@

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@cfa69] and multidimensional scaling (MDS) [@borg2005]. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)
<!--### Notation -->

>>>>

bibliography: paper.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
---

@@

<!-- - Connection between NLDR and MDS-->
Historically, \kD{}   representations of \pD{} data have been computed using multidimensional scaling (MDS) [@kruskal1964], which includes principal components analysis (PCA) [@jolliffe2011] as a special case. (A contemporary comprehensive guide to MDS can be found in @borg2005.) The \kD{}   representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{}   layout is constructed by minimizing a stress function that differences distances between points in \pD{}   with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

@@

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@cfa69] and MDS. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)
<!--### Notation -->
